\texttt{dispRity} manual

### 2015-11-29 doc/dispRity-manual.Rnw
added random seed to manual
Updated manual

\section{Before starting}
\subsection{Glossary}
Because this package is aimed to be multidisciplinary, many names or terms used in this tutorial might be non-familiar to certain fields.
Here is a list of what are the exact meaning of these term:
\begin{itemize}
\item \textbf{Ordinated space}: it designates the mathematical multidimensional object studied here. In morphometrics, this one is often referred as being the morphospace, or the cladisto-space for cladisticts or the eco-space for ecology, etc. In practice though, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often $>$ 3!) and the rows represent the elements within this space.
\item \textbf{Elements}: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
\item \textbf{Dimensions}: it designates the columns of the ordinated space. The dimensions can also be referred to as axis.
\item \textbf{Series}: it designates sub-samples of the ordinated space. Basically a series contain the same number of dimensions as the morphospace but might contain a smaller number of elements. For example, if our ordinated space is composed of birds and mammals (i.e. the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but the same 50 dimensions) to look at the difference in disparity between both groups.
\end{itemize}

@@

<<installation, eval=FALSE>>=
install.packages("devtools")
library(devtools)
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
@

\noindent Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
For the piping-hot (but potentially full of bugs) version, one can change the argument \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.

\subsection{Data}
In this tutorial we are going to use a subset of the ordinated cladistic data from \cite{beckancient2014} that contains 50 taxa ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
Please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes covering some specific example of the package with ecological data.

@@

\subsection{A quick go through}
Here is a really crude and quick go through the package to show some of the features of this package.
Note that all these features will be discussed in more details below.

<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)), FADLAD = BeckLee_ages)
## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)
## Calculating disparity
sum_of_variances <- dispRity(bootstrapped_data, metric = c(sum, variances))
## Summarising the results
summary(sum_of_variances)
plot(sum_of_variances)
## Testing some the effect of time on disparity
summary(test.dispRity(sum_of_variances, test = aov, comparisons = "all"))
@

@@

<<dispRity class1, eval=FALSE>>=
str(sum_of_variances)
## That's a more than 4500 lines of output!
@

@@

<<dispRity class2>>=
## Which class is the sum_of_variances object?
class(sum_of_variances)
## What's in the object
names(sum_of_variances)
## We can summarise it using the S3 method print.dispRity
sum_of_variances
## This displays some information about what's in the object
@

@@

<<dispRity class3, eval=FALSE>>=
## Displaying the full object
print(sum_of_variances, all=TRUE)
@

@@

<<dispRity class4>>=
## Extracting some specific series from the disparity object
series_1_and_4 <- get.dispRity(sum_of_variances, what=c(1,4))
series_1_and_4
## The observed disparity
extract.dispRity(sum_of_variances)
## The list of bootstrapped scores of disparity
str(extract.dispRity(sum_of_variances, observed = FALSE))
@

@@

<<dispRity modular>>=
## Disparity measured as the mean
summary(dispRity(sliced_data, metric = mean))
## A function for measuring the sum divided by the length (the mean!)
sum.by.length <- function(X)  sum(X)/length(X)
## Disparity measured as the mean divided by the length
summary(dispRity(sliced_data, metric = sum.by.length))
## Giving the exact same results!
@

@@

\section{Functions}
One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter \textit{series}).
These series correspond at regions of the ordinated space that we want to compare to each other.

@@

## This object contains three elements
names(crown_stem)
## With "data" being the list sub-matrices
str(crown_stem$data)
## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)
## and finally, "series" containing information on the series type (custom) and names
## (crown and stem)
crown_stem$series
@

@@

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The \texttt{time.series} function allows to create series that contain all the elements present at specific points in time.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time series can be performed by using the \texttt{method} option:
\begin{enumerate}
\item discrete time series (or time-binning) using \texttt{method = "discrete"};
\item continuous time series (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}
For the time-slicing method details see \cite{GuillermeSTD}.
For both methods, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
\begin{enumerate}
\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
\item defining the time-slicing places (when using \texttt{method = "continuous"}).
\end{enumerate}
Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a data frame containing the First/Last Occurrence Data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.

@@

<<discrete.series2>>=
## Automatically generate three equal length bins:
time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete", time = 3)
@
\noindent We now have three time bins of 44.50368 Ma each.

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the tree length as long as at least three elements are present at that point in time (an extra 1\% is added until reaching the required minimum of three elements).
We can also ask to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.

@@

<<continuous.series>>=
## Generating four time slices every 40 Ma assuming a gradual evolution model
time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
time_slices
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
@

@@

Once we obtain our different series, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction.
The number of bootstrap pseudo-replicates can be defined using the \texttt{bootstraps} option (see below)
.
<<boot.matrix1>>=
boot.matrix(data = BeckLee_mat50)
## As we can see, the output is also a dispRity object that is summarised
## automatically, and gives information on the data as well as the number of
## bootstraps and the bootstraps methods.
@

Additionally, this function allows to control the bootstrap algorithm through the \texttt{boot.type} argument.
Currently two algorithms are implemented:
\begin{enumerate}
\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
\item \texttt{"single"} where only one random elements is replaced by one other random elements each pseudo-replication
\end{enumerate}
This function also also to rarefy the data using the \texttt{rarefaction} argument.
The default argument is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e. remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
Finally, one last argument, \texttt{rm.last.axis} allows to remove a certain amount of dimensions (or axis) for the ordinated space.
This can be logical argument where \texttt{FALSE} (default) will not remove any dimension and \texttt{TRUE} will remove the last dimensions that bear up to 5\% of the total ordinated space's variance.

@@

Of course, one could be interested in directly supplying the subsampled matrices generated above directly to this function.
In fact, it can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.
let's bootstrap our previously generate series:

<<boot.matrix3>>=
## Bootstrap and full rarefaction on the crown/stem series
crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
## Bootstrap on the time binning/slicing series
time_binsBS <- boot.matrix(time_bins)
time_slicesBS <- boot.matrix(time_slices)
## Note that all these objects are of class dispRity
crown_stemBS
time_binsBS
time_slicesBS
@

@@

Now we're going to see the functionalities of the core function of this package: the \dispRity function.
This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.

Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
%To change in future version!
The disparity metric boils down to a single value that describes the matrix.
Some are classic like the sum or the product of the ranges or the variances of the matrix columns (i.e. the any-o-space axis/dimensions) \cite{Wills1994}.
However, many more exist like the median distances between the taxa and the any-o-space centroid \cite{GuillermeSTD}.

One can usually decompose the disparity metrics into two elements:
\begin{enumerate}
\item the \textbf{class metric} that is a descriptor of the matrix. For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
\end{enumerate}
Basically the combination can be infinite between the class and summary metrics.
For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others (see section \ref{whatsleft})...

In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
It is even possible to use your very own class and summary metrics!
This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version (see \ref{whatsleft}).

To use these metrics pairs in the \dispRity function, it's pretty easy:

<<dispRity1>>=
dispRity(BeckLee_mat50, metric = c(sum, ranges))
## Yep, that's it. The order of the metrics does not matter (the function detects
## automatically which one is the class and the summary metric) so the combinations
## are numerous!
dispRity(BeckLee_mat50, metric = c(sum, variances))
dispRity(BeckLee_mat50, metric = c(prod, centroids))
dispRity(BeckLee_mat50, metric = c(mode.val, ranges))
## Or funny user defined ones!
total.range <- function(X) abs(range(X)[1]-range(X)[2])
dispRity(BeckLee_mat50, metric = c(total.range, ranges))
## That would be disparity as the range of the axis ranges!
@

Of course, this function can take a simple ordinated matrix but, more interesting it can also deal with a list of matrices (from the \texttt{cust.series} or \texttt{time.series} functions for example) or a bootstrapped or/and rarefied matrix (from the \texttt{boot.matrix} function for example).

<<dispRity2>>=
## For example, let's calculate the sum of variances from our crown/stem mammals
dispRity(crown_stem, metric = c(sum, variances))
## Or even better, from the bootstrapped data
disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, variances))
## Note that the bootstrapped + rarefied data takes some time to run (it's
## calculating disparity from 46 matrices!)

## Let's calculate a different metric for the disparity-through-time analysis
disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
@

Again, note that the results are \dispRity objects and can displayed through the S3 \texttt{print.dispRity} method to remind users what are in these objects.
Also note that an even faster parallel version will be developed in the future (see section \ref{whatsleft}).

@@

As shown in these examples above however, even though the disparity has been calculated in many way, the function doesn't directly outputs the results, rather just a summary of the analysis.
To really display the results we can use the S3 \texttt{summary.dispRity} and \texttt{plot.dispRity} functions.

\subsubsection{\texttt{summary} function}
This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.

>>>>

\section{Before starting}
\subsection{Glossary}
Because this package is aimed to be multidisciplinary, many names or terms used in this tutorial might be non-familiar to certain fields.
Here is a list of what are the exact meaning of these term:
\begin{itemize}
\item \textbf{Ordinated space}: it designates the mathematical multidimensional object studied here. In morphometrics, this one is often referred as being the morphospace. However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc. In practice, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often -- but not necessarily -- $>$ 3!) and the rows represent the elements within this space.
\item \textbf{Elements}: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...
\item \textbf{Dimensions}: it designates the columns of the ordinated space. The dimensions can also be referred to as axis.
\item \textbf{Series}: it designates sub-samples of the ordinated space. Basically a series contain the same number of dimensions as the morphospace but might contain a smaller number of elements. For example, if our ordinated space is composed of birds and mammals (i.e. the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but the same 50 dimensions) to look at the difference in disparity between both groups.
\end{itemize}

@@

<<installation, eval=FALSE>>=
if(!require(devtools)) install.packages("devtools")
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
@

\noindent Note that we use the \texttt{release} branch here which is version \Sexpr{packageVersion("dispRity")}.
For the piping-hot (but potentially full of bugs) version, you can change the argument \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.

\subsection{Data}
In this tutorial we are going to use a subset of the ordinated cladistic data from \cite{beckancient2014} that contains 50 taxa (elements) ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
Please refer to the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} for other vignettes covering some specific example.

@@

\subsection{A quick go through}
Here is a really crude and quick analysis to go through the package, showing some of its features.
Note that all these features will be discussed in more details below.

<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
    FADLAD = BeckLee_ages)
## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)
## Calculating disparity
sum_of_variances <- dispRity(bootstrapped_data, metric = c(sum, variances))
## Summarising the results
summary(sum_of_variances)
plot(sum_of_variances)
## Testing the effect of time on disparity
summary(test.dispRity(sum_of_variances, test = aov, comparisons = "all"))
@

@@

<<dispRity class1, eval=FALSE>>=
str(sum_of_variances)
## That's a more than 4000 lines of output!
@

@@

<<dispRity class2>>=
## What is the class of the sum_of_variances object?
class(sum_of_variances)
## What is in the object?
names(sum_of_variances)
## Summarising it using the S3 method print.dispRity:
sum_of_variances
## This displays some basic information about the object content
@

@@

<<dispRity class3, eval=FALSE>>=
## Displaying the full object
print(sum_of_variances, all = TRUE)
@

@@

<<dispRity class4>>=
## Extracting some specific series from the disparity object
series_1_and_4 <- get.dispRity(sum_of_variances, what = c(1,4))
series_1_and_4
## The observed disparity
extract.dispRity(sum_of_variances)
## The list of bootstrapped scores of disparity
str(extract.dispRity(sum_of_variances, observed = FALSE))
@

@@

<<dispRity modular>>=
## Disparity measured as the mean value of the ordinated matrix
summary(dispRity(sliced_data, metric = mean))
## A function for measuring the sum divided by the length (the mean!)
sum.divide.by.length <- function(X) sum(X)/length(X)
## Disparity measured as the mean divided by the length
summary(dispRity(sliced_data, metric = sum.divide.by.length))
## Giving the exact same results!
@

@@

\section{Functions}
One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter \textit{series}).
These series correspond to regions of the ordinated space that we want to compare to each other.

@@

## This object contains three elements
names(crown_stem)
## With "data" being the list sub-matrices
str(crown_stem$data)
## "elements" being the list of taxa in the original ordinated matrix
str(crown_stem$elements)
## and finally, "series" containing information on the series type (custom)
## and names (crown and stem)
crown_stem$series
@

@@

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The \texttt{time.series} function allows to create series that contain all the elements present at specific points or during specific periods in time.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time series can be performed by using the \texttt{method} option:
\begin{enumerate}
\item discrete time series (or time-binning) using \texttt{method = "discrete"};
\item continuous time series (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}
For the time-slicing method details see \cite{GuillermeSTD}.
For both methods, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
\begin{enumerate}
\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
\item defining the time slices (when using \texttt{method = "continuous"}).
\end{enumerate}
Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a data frame containing the First/Last Occurrence Data (\texttt{FAD/LAD}) for taxa that span over a longer time than the tips/nodes age.

@@

<<discrete.series2>>=
## Automatically generate three equal length bins:
time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = 3)
@
\noindent We now have three time bins of 44.50368 million years each.

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1\% of the tree length as long as at least three elements are present at that point in time.
The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
It is also possible to include nodes in each bin by using \texttt{inc.nodes = TRUE} and providing a matrix that contains the ordinated distance between tips \textit{AND} nodes.

@@

<<continuous.series>>=
## Generating four time slices every 40 million years assuming a gradual
## evolution model
time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
time_slices
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
@

@@

Once we obtain our different series, we might want to bootstrap and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
The \texttt{boot.matrix} allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction.
The number of bootstrap pseudo-replicates can be defined using the \texttt{bootstraps} option (see below).

<<boot.matrix1>>=
boot.matrix(data = BeckLee_mat50)
## As we can see, the output is also a dispRity object that is summarised
## automatically, and gives information on the data as well as the number of
## bootstraps and the bootstraps methods.
@

Additionally, this function allows to control the bootstrap algorithm through the \texttt{boot.type} argument.
Currently two algorithms are implemented:
\begin{enumerate}
\item \texttt{"full"} where the bootstrapping is entirely stochastic (all the data is bootstrapped)
\item \texttt{"single"} where only one random elements is replaced by one other random elements each pseudo-replication
\end{enumerate}

This function also allows to rarefy the data using the \texttt{rarefaction} argument.
The default argument is \texttt{FALSE} but it can be set to \texttt{TRUE} to fully rarefy the data (i.e. remove $n$ elements for the number of pseudo-replicates, where $n$ varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
It can also be set to a fix \texttt{numeric} value (or a set of numeric values).
Finally, one last argument, \texttt{rm.last.axis} allows to remove a certain amount of dimensions (or axis) for the ordinated space.
This can be logical argument where \texttt{FALSE} (default) will not remove any dimension and \texttt{TRUE} will remove the last dimensions that bear up to 5\% of the total ordinated space's variance.

@@

Of course, one could be interested in directly supplying the sub-sampled matrices generated above directly to this function.
In fact, it can also deal with a list of matrices or with a \dispRity object output from the \texttt{cust.series} or \texttt{time.series} functions.

<<boot.matrix3>>=
## Bootstrap and full rarefaction on the crown/stem series
crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
## Bootstrap on the time binning/slicing series
time_binsBS <- boot.matrix(time_bins, bootstraps = 100)
time_slicesBS <- boot.matrix(time_slices, bootstraps = 100)
## Note that all these objects are of class dispRity
crown_stemBS
time_binsBS
time_slicesBS
@

@@

This function is a modular function that allow to simply (and quickly!) calculate disparity from a matrix.

Because disparity can be measured in many ways, this function is a tool to measure disparity \textit{as defined by the user}.
In fact, the \dispRity function intakes two main arguments: the data and the disparity metric.
The disparity metric is a function or a set of functions that summarises the ordinated matrix to a single value that represents, in our example, the diversity of morphologies.

The \dispRity algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
For more details on this algorithm please refer to the \href{https://github.com/TGuillerme/dispRity}{metric vignette}.

In practice, the \dispRity function intakes one or more functions as a definition of disparity.
Several of these functions will be already implemented in other packages (such as \texttt{stats::median}, \texttt{base::sum}, etc.); some others are implemented in this package (listed in \texttt{?dispRity.metric}) and finally some others will be defined by the users.
The \texttt{make.metric} function is designed to help users create and test their own disparity metric functions.
In practice, the use of these metrics in the \dispRity function, is pretty easy:

<<dispRity1>>=
dispRity(BeckLee_mat50, metric = mean)
## This defines disparity as the mean value of the morphospace (using
## mean::base).

## It is also possible to combine multiple functions:
dispRity(BeckLee_mat50, metric = c(sum, variances))
## Defining disparity as the sum (sum::base) of the variances
## (variances::dispRity) of each dimension of the morphospace.

dispRity(BeckLee_mat50, metric = c(prod, centroids))
## For the product (prod::base) of the distances between each elements and the
## centroid of the morphospace (centroids::dispRity).

## Or user defined ones:
total.range <- function(X) abs(range(X)[1]-range(X)[2])
dispRity(BeckLee_mat50, metric = c(total.range, centroids))
## For the range (total.range, user defined) of the distances between each
## elements and the centroid of the morphospace (centroids::dispRity).

## Or more complex ones:
vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
## For the standard deviation (sd::stats) of the variance (variances::dispRity)
## of each column of the variance/covariance matrix (var::stats).

## Note that the order of the function is not important since the levels of
## each function are automatically detected by the dispRity algorithm.
vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
all(summary(vars) == summary(vars2))
@

\noindent Note that these functions do not directly output the disparity values but only the summary of the \dispRity objects.
To display the results, see the section \ref{summary}.

In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the \dispRity objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
In this example we are going to define disparity as being the sum of the variances in each dimension)=.

<<dispRity2>>=
## Disparity in crown and stem mammals:
disp_crown_stemBS <- dispRity(crown_stemBS, metric = c(sum, variances))

## Disparity through time:
disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
@

\noindent Note that the computational time was longer for the rarefied data (\texttt{crown\_stemBS}) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
A faster parallel version will be implemented in future releases (see section \ref{whatsleft}).

% STOPPED HERE

@@

As shown in these examples above however, even though the disparity has been calculated in many way, the function doesn't directly outputs the results, rather just a summary of the analysis.

This function intakes a \dispRity object plus various options namely the \texttt{quantile} values for the confidence intervals levels; the \texttt{cent.tend} for the central tendency to use for summarising the results and two \textit{visual} options which are whether to recall the \dispRity options and how much digits are wanted in the results.
